Generative Dynamics of Supreme Court Citations: Analysis with a New Statistical Network Model

### 2020-08-31 Tex/PA_rr/main.tex
meshed ideological explanations

We offer three contributions . First, we advance our understanding of the factors that drive citations between U.S. Supreme Court opinions. Second, for those who study judicial citations in general (e.g., U.S. state supreme courts \citep{hinkle2016transmission}, international courts \citep{lupu2012precedent},  German lower courts \citep{berlemann2020disposition}), we illustrate network-theoretic considerations that are likely to apply beyond the context of Supreme Court opinions. Third, we offer a novel extension of a statistical model for networks, and disseminate this model as a package for the R statistical software \citep{cergm}, which can be used for any form of citations (e.g., patent citations are used to measure both causes and consequences of innovation in the field of political economy \citep{akcigit2018growth,dincer2019does},  citations among academic journal articles have been used to study gender bias in political methodology \citep{dion2018gendered} and international relations \citep{maliniak2013gender},  and citations in documents produced in the policymaking process have been used to study links between public policy and scientific expertise \citep{costa2016science,koontz2018use}). Our contributions are both substantive and methodological. We focus most heavily on the first---the study of the U.S. Supreme Court citation network, since we see it as the most substantial innovation with respect to the existing literature that we offer in the current paper.

@@

These three properties---transitivity, reciprocity, and popularity form the core of our network theory of U.S. Supreme Court citations. We also seek to model the effects of covariates (i.e., case features) on citation formation.  Next we introduce a modeling framework that can incorporate all of these effects into a single model, and adapt to the structural constraints of citation networks.

@@

\subsubsection{Covariate Terms}
Effects of the exogenous covariates included in the c-ERGM are presented in Figure \ref{SCC_results_2}. Most of the exogenous covariate effects align with expectations and have relatively stable parameter estimates over time. We find that cases are more likely to cite those that (1) have been decided most recently, (2) were authored by the same justice, (3) are in the same issue area, and (4) have not been overruled. Surprisingly, we do not find consistent effects for majority coalition size or for the two covariates that involve justice ideology. The effect of majority coalition size was generally not statistically significant until the later portion of the period examined; ideological distance between two cases had signs and significance levels that are highly inconsistent; and the effect of the ideological breadth of the potentially cited case was effectively zero for the entire period. A purely measurement-based explanation of our findings is that we need a more precise measure of the ideological position of an opinion, as far as citation behavior goes, than the median ideal point of the justice in the majority coalition. Clustering that results from ideological homophily that is not effectively accounted for by our measure would be picked up, in part, by our measures of transitivity, which are statistically and substantively significant throughout the period studied. However, we do observe connections between our findings and the literature on decision-making on the Supreme Court.

Our results leave open the question of whether and how citations are shaped by ideological factors. While we did not formally hypothesize about temporal dynamics in the effects, a closer look at the observed trends allows us to identify some patterns that suggest directions for future research in this area. In this discussion, we focus on the shift in the effect of ideological distance beginning with the Rehnquist court and extending into the Roberts court.

>>>>

We offer three contributions. First, we advance our understanding of the factors that drive citations between U.S. Supreme Court opinions. Second, for those who study judicial citations in general (e.g., U.S. state supreme courts \citep{hinkle2016transmission}, international courts \citep{lupu2012precedent},  German lower courts \citep{berlemann2020disposition}), we illustrate network-theoretic considerations that are likely to apply beyond the context of Supreme Court opinions. Third, we offer a novel extension of a statistical model for networks, and disseminate this model as a package for the R statistical software \citep{cergm}, which can be used for any form of citations. For example, patent citations are used to measure both causes and consequences of innovation in the field of political economy \citep{akcigit2018growth,dincer2019does}, citations among academic journal articles have been used to study gender bias in political methodology \citep{dion2018gendered} and international relations \citep{maliniak2013gender}, and citations in documents produced in the policymaking process have been used to study links between public policy and scientific expertise \citep{costa2016science,koontz2018use}. Our contributions are both substantive and methodological. We focus most heavily on the first---the study of the U.S. Supreme Court citation network, since we see it as the most substantial innovation with respect to the existing literature that we offer in the current paper.

@@

These three properties---transitivity, reciprocity, and popularity---form the core of our network theory of U.S. Supreme Court citations. We also seek to model the effects of covariates (i.e., case features) on citation formation.  Next we introduce a modeling framework that can incorporate all of these effects into a single model, and adapt to the structural constraints of citation networks.

@@

\subsubsection{Covariate Terms}
Effects of the exogenous covariates included in the c-ERGM are presented in Figure \ref{SCC_results_2}. Most of the exogenous covariate effects align with expectations and have relatively stable parameter estimates over time. We find that cases are more likely to cite those that (1) have been decided most recently, (2) were authored by the same justice, (3) are in the same issue area, and (4) have not been overruled. Surprisingly, we do not find consistent effects for majority coalition size or for the two covariates that involve justice ideology. The effect of majority coalition size was generally not statistically significant until the later portion of the period examined; ideological distance between two cases had signs and significance levels that are highly inconsistent; and the effect of the ideological breadth of the potentially cited case was effectively zero for the entire period.

Our results leave open the question of whether and how citations are shaped by ideological factors. A purely measurement-based explanation of our findings is that we need a more precise measure of the ideological position of an opinion, as far as citation behavior goes, than the median ideal point of the justice in the majority coalition. Clustering that results from ideological homophily that is not effectively accounted for by our measure would be picked up, in part, by our measures of transitivity, which are statistically and substantively significant throughout the period studied.

Beyond potential measurement issues, we do observe connections between our findings and the literature on decision-making on the Supreme Court. These connections are evident from a closer look at temporal dynamics in the effects. In this discussion, we focus on the shift in the effect of ideological distance beginning with the Rehnquist court and extending into the Roberts court.
